A somehow subtle pitfall is related to the begin-of-line pre-condition.
When using the '\verb|^|' sign at the beginning of a line it is tempting to
say "do not worry, this matches at {\it any} begin of line"---well it
does not. The important thing to understand is that it matches when the lexical analysis
    step {\it starts} at the beginning of a line. The alarm signal should ring
   if begin-of-line is triggered and a whitespace pattern is defined that includes newline.
Consider the following patterns being defined:

\begin{lstlisting}
define {
   WHITESPACE    [ \t\n]+
       ....
   GREETING       ^[ \t]*hello
}

mode x_y_z : {
    {WHITESPACE}  => TKN_WHITESPACE;
    {GREETING}    => TKN_GREETING(Lexeme);
}

\end{lstlisting}

Where the {\tt hello} greeting is to be matched after newline and some possible whitespace.
Now, given the character stream:

\begin{lstlisting}
   ...
   hello
\end{lstlisting}

will {\it not} send a GREETING token. This is because the whitespace pattern eats
the newline before the 'hello-line' and even the whitespace before hello. Now, the
next analysis step starts right before {\tt hello} and this is not the beginning
of a line. Splitting the whitespace eater into newline and non-newline helps:

\begin{lstlisting}
define {
   WHITESPACE_1  [ \t]+
   WHITESPACE_2  [\n]+
   GREETING      ^[ \t]*hello
}

mode x_y_z : {
    {WHITESPACE_1}  => TKN_WHITESPACE;
    {WHITESPACE_2}  => TKN_WHITESPACE;
    {GREETING}      => TKN_GREETING(Lexeme);
}

\end{lstlisting}

Now, the first whitespace eating ends right before newline, then the second
whitespace eating ends after newline, and the hello-greeting at the beginning
of the line can be matched.

Another pitfall is, again, related to precedence. Make sure that if there
are two patterns with the same core pattern {\tt R}, then the pre- or post-conditioned patterns
must be defined {\it before} the un-conditioned pattern. Otherwise, the
pre- or post-conditioned patterns may never match. Recall section
\ref{formal/patterns/context-free-pitfalls} for a detailed discussion on precedence pitfalls.

